Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-09 ResponseToReviewers/ResponseToReviewers.tex

\documentclass[10pt,a4paper]{letter}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}

\begin{document}

\signature{Trever T. Hines}

\begin{letter}{}
\opening{Dear Editor,}

We were pleased with the positive comments from you and the reviewers, and we feel that the comments brought up valid concerns. We hope that we have addressed these concerns sufficiently in our revision. In order to fully address the comments, we have included a supplementary material document.  We have also made some changes to the manuscript which were not specifically requested by the reviewers.  Most of these changes were stylistic and do not change the content of the manuscript.  One of the more substantial changes was that we merged Figures 7 and 14, which are both showing observed and predicted radial components of postseismic displacements, facilitating direct comparison of the figures.

We added supporting figures S3 and S4, which break down the predicted displacements of our preferred model into an elastic and viscoelastic component.  We feel that these figures help support our discussion on lines 509 to 523, where we attribute inferences of fault slip and viscoelastic relaxation to different aspects of the observed postseismic displacements.

On line 500 to 502 of our previous version of this manuscript we make the claim that “After one year, afterslip is inferred to be deeper down on the Sierra Cucapah segment, which is describing much of the sustained near-field postseismic deformation.”  This sentence is not accurate of the preferred model we present in the manuscript. As can be clearly seen in the new figures S3-S4, afterslip is describing some of the later near-field deformation, although most of it is being described by viscoelastic relaxation.  We have reworded this passage and updated the introduction and conclusion to accurately describe our interpretations of the mechanisms driving later near-field deformation.  This does not significantly alter the main conclusions of our manuscript, because we are very clear, both in the original and revised versions, that the deformation mechanism driving later near-field deformation cannot be definitively resolved with these data.

Each of the reviewer comments are included below followed by our response and a description of how these comments were addressed in this revised manuscript.

%% REVIEWER 1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textbf{Reviewer \#1:}\newline

\textbf{Review of the paper : Revealing transient strain in geodetic data
with Gaussian process regression}

\textbf{Overview of the work: The paper focuses on a new non-parametric
Bayesian method to estimate transient strain rate from GNSS data. It
is based on Gaussian Process regression. The method also includes an
outlier removal processing step. This technique is then applied to the
detection of slow slip events in Cascadia. The validity of the
detected transient signals is accomplished via comparison with the
seismic tremor recorded along the Cascadia range.}

\textbf{The article is well-written and your algorithm is very interesting. I
think the use of Gaussian Process Regression is novel and open new
roads in GNSS data analysis. I believe that you spend a lot of efforts
in trying to justify most of the assumptions behind your model,
together with the effects of using different kind of prior. Perhaps,
that overshadows slightly your results. You could consider to
reorganize the article to make it more readable for geophysicists.}

\textbf{I also require that a comparison between your outlier removal model
and one other algorithm could help to support your claims about the
efficiency of your method.}

\textbf{Otherwise, I have underlined a few clarifications needed about the
stochastic model and some paragraphs within the text.}

\textbf{General Comments:}

\textbf{About the method:}

\textbf{The method is well established. The assumptions to justify the various
hypothesis on the deterministic models of the GNSS time series are
generally well supported (seasonal signal, tectonic rate, offsets).}

\textbf{1- Now, the correlation of the stochastic processes is not completely
clear. When the authors justify the model behind equation 3, wij is a
normally distributed, uncorrelated noise. In statistic, it is equal to
a WGN process. Then, the authors introduce a separate parameter (eta)
which models the temporally correlated noise. This parameter is also
following a Gaussian distribution, with a zero-mean and a covariance
matrix $C_{\eta}$ (i.e. line 10 p5). This approach is not common in
modelling GNSS time series. The standard is based on Williams 2003,
where the author established a stochastic model with a covariance
matrix as sum of identity matrix for the white noise and another
matrix which represents the coloured noise. The coloured noise
covariance matrix is defined differently (see the literature – Bock and
Melgar Reviews of Geophysics 2016, He et al., Journal of geodynamics
2017).}

\textbf{2- First, the author should give a summary of stochastic noise
modelling in GNSS. I also think that a discussion is needed to relate
to previous models. Why using separated stochastic models? Also I
would comment on the possibility that the (low) spatial correlation
between the parameters in your deterministic model (i.e. secular
velocities) could be absorbed in the estimation of the covariance
matrix $C_{\eta}$.}

\textbf{Now, I am confused When looking at p.8 Section 4.1, the description of
the noise models should be given as a subsection of Section 2 p.4. I
would also improve the literature review. Several other models have
also been discussed (i.e. Montillet et al., 2014 uses a fractional
Brownian motion model; ARMA, ARFIMA, GGM or Band pass noise – see Het
et al., 2017).}

\textbf{Outlier detection:}

\textbf{1- P7 : I would give a longer summary in the introduction about all
the efforts in outlier detection in GNSS time series. The first
paragraph p.7 is not enough.}

\textbf{2- Also, the paper would be improved if the proposed outlier detection
method is compared with existing ones such as the Hector software
package (Bos et al., 2013). For example, the comparison between this
method and another one could be done when applying the algorithm in p.
13 (last paragraph below equation 18).}

\textbf{Results:}

\textbf{The results are generally well explained. However, the use of
different priors and the comparison in the text is sometimes
confusing. Perhaps, a table could summary the different priors and the
main results. It would ease the reading of Section 4.2, 4.3 and then
the discussion in Section 5.}

\textbf{Minor issues:}

\textbf{P2 line 30 : “fidelity” replace with “reliability”.}

\textbf{P2 line 30-31 “ Developing and improving upon methods for deriving
secular...area of research.” Need references.}

\textbf{p.2 line 43 :”too large of an area” ... perhaps “ a very large area”}

\textbf{P5 line 32 : “formal data uncertainty”: can you define it?}

\textbf{P 6 : Need reference to show Equation 15.}

\textbf{P 7 line 54-60: This paragraph can be further enhanced with references
(Bock and Melgar, 2016, Gazeaux et al., 2013).}

\textbf{P 8 : first paragraph. add references on Cascadia range (Aguiar,
Melbourne, Scriver 2009), (Melbourne and Szeliga 2005), (Melbourne and
Webb, 2003, 2002).}

\textbf{P.8 line 20 delete www.unavco.org (only in Acknowledgment)}

\textbf{p. 11 line 45 “common mode error” you need to add some references and
define it properly.}

\textbf{p.13 : “Wendland functions have compact support and hence their
corresponding covariance.. sparse”. I would just mention that the
associated covariance matrices are sparse.}

\textbf{p.13 line 22 what is an “isotropic Gaussian process”? please add a
definition.}

\textbf{P.20 : “ Using a compact covariance function ..” Again I would just
say that the covariance matrix is sparse and that s why you have the
mathematical simplification.}

\textbf{p.20 line 33 “ computational burden” I think that refers to p.20 line
60 “prohibitive when using sveral years of daily GNSS”. It would
interesting to have some number. How do you quantify the computation
time to process longer and longer time series?}

%% REVIEWER 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textbf{Reviewer \#2:}\newline

\textbf{Comments to the Author(s)}
\textbf{Hines/Hetland GJI Review 08/2017}

\textbf{This is a well written paper and I have just a few minor
clarifications and suggestions below.  One point that might be of
interest to address is would it be possible to “approximate” this
approach so that it can run much faster?  I gather from the text that
the method is intensive for CPU and memory use.}

\textbf{Page 3: Last line: min(t,t’) needs as scaling factor to make units for
covariance e.g., m$^2$/yr}

\textbf{Page 4: Stochastic model: It would seem that the assumption of no
covariance between the north and east components is not a very good
one since there is a tectonic framework in which the transients occur
and its unlikely this framework aligns along the Cardinal directions.
Maybe add a statement of impact of the neglect and possibly the idea
of reorienting the “axes” to align with the tectonic framework e.g.,
perpendicular to the subduction zone interface in this case?}

\textbf{Page 4/5 Equation (3): Is it worth discussing at this point whether
the estimates of u and n (eta) can be separated without explicit prior
knowledge (which we may not have).  For example, orbit modeling errors
on one satellite (e.g., due to an unknown yaw problem in the
spacecraft) wilt generate a spatially and temporally correlated error
in $d_{ij}$.  How does this (stochastically non-modeled) error not project
in the u estimates?}

\textbf{Page 6: Equation 10: Maybe it is discussed later in the paper but it
is probably worthwhile stating at this point how the inverse is
performed with a block 0 in the matrix?}

\textbf{Page 6: Equation 12: Maybe some additional explanation is needed here.
How are these partial derivatives formed.  My understanding the u
estimates at this point correspond one-to-one with the positions and
times of the GNSS position determinations.  Did I miss something?}

\textbf{Page 8: Partly addresses issue raised about eqn 3.  It probably should
be noted that wether conditions east of -121 deg longitude are
different to the coast so there could be possible problems with this
noise model.}

%% Editor Review
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\textbf{Editor:}\newline

\textbf{Editor’s Review of GJI MS 17.0593 ‘‘Revealing transient strain...’’ by
Hines and Hetland This is an novel algorithm applied to a problem of
great current interest. Based on the reviewer’s comments and my own
reading, I am recommending moderate revision. The authors’ response
should include a description of their changes, and also a version of
the paper on which these can be seen (just as highlights in the text).}

\textbf{Many of the reviewers’ comments address places where the paper could
be more clearly written, and my overall comment would be the same. The
current version perhaps spends a little too much time on the
background and not enough on explaining the method and its derivation.
As shown by the references below (these include some the authors
cite), and the Dermanis paper, which I have also attached, strain
estimation from scattered data is an area in which there has been a
lot more prior work than the authors mention. I am including these
references not because the authors need to cite many or even most of
them, but only to make the point that the problem of estimating
strain, and of looking for transients, is a familiar one: for many
readers, myself among them, much more familiar than the statistical
methods applied here. So a slightly fuller discussion of these methods
(perhaps a little more background and some reminders of, eg, what a
hyperparameter is) would be welcome to many readers; and increased
clarity of presentation usu- ally means that the paper will be used
and cited more.}

\textbf{I feel the same way about the section on detecting outliers: again, a
well-worked area. Iter- ative fitting and outlier detection, as an
overall strategy, hav e been used for a long time. I appreci- ate that
by imposing spatial and temporal conditions you can identify outliers
not otherwise obvi- ous, but it would help to show an example of this,
rather than the current Figure 4, in which most of the identified
cases are ones that any method could easily find.}

\textbf{A few specifics (numbers are page+line):}

\textbf{2+26: possibility not risk: the latter is a technical term in seismic
hazard.}

\textbf{2+47-48: I do not think this description of Shen’s method (which I’ve
used and programmed) is right, since it allows the deformation
gradients to vary in space. I’ve always thought of it as a kind of
adaptive kernel smoother, somewhat like applying 2-D loess; so it is
only depen- dent on nearby points, and has no assumption of uniformity
anywhere. (Actually this raises a more general question: is the method
given here use local or global support?)}

\textbf{3+7: could also}

\textbf{3+32: redo the sentence; as it stands, you are saying that the SCEC
exercise calculates strain rates.}

\textbf{3+37: a geophysical signal}

\textbf{3+47: function not signal}

\textbf{3+56: by a Gaussian, not with}

\textbf{4+37: displacement (not plural)}

\textbf{6+56: I suspect that equation (15) took some effort to derive; could
some details go into an Appendix?}

\textbf{Figure 6: I like that the strain uncertainties can be shown, but is
there a reason not to plot principle strains using arrows, showing
errors by an ellipse around the tips?}

\closing{Thank you for your continued consideration,}

\end{letter}
\end{document}

### 2017-11-09 ResponseToReviewers/ResponseToReviewers.tex

\textbf{Review of the paper : Revealing transient strain in geodetic data
with Gaussian process regression}

\textbf{Overview of the work: The paper focuses on a new non-parametric
Bayesian method to estimate transient strain rate from GNSS data. It
is based on Gaussian Process regression. The method also includes an
outlier removal processing step. This technique is then applied to the
detection of slow slip events in Cascadia. The validity of the
detected transient signals is accomplished via comparison with the
seismic tremor recorded along the Cascadia range.}

\textbf{The article is well-written and your algorithm is very interesting. I
think the use of Gaussian Process Regression is novel and open new
roads in GNSS data analysis. I believe that you spend a lot of efforts
in trying to justify most of the assumptions behind your model,
together with the effects of using different kind of prior. Perhaps,
that overshadows slightly your results. You could consider to
reorganize the article to make it more readable for geophysicists.}

I have rewritten the methods section to contain more examples and I
elaborated on concepts that may not be familiar to a reader that does
not have a strong background in stochastic processes. I sincerely hope
that my revised draft is more readable to geophysicists.

@@

\textbf{I also require that a comparison between your outlier removal model
and one other algorithm could help to support your claims about the
efficiency of your method.}

(See my response to the comment ``Also, the paper would be improved if
the proposed outlier detection ...")

@@

\textbf{About the method:}

\textbf{The method is well established. The assumptions to justify the various
hypothesis on the deterministic models of the GNSS time series are
generally well supported (seasonal signal, tectonic rate, offsets).}

\textbf{1- Now, the correlation of the stochastic processes is not completely
clear. When the authors justify the model behind equation 3, wij is a
normally distributed, uncorrelated noise. In statistic, it is equal to
a WGN process. Then, the authors introduce a separate parameter (eta)
which models the temporally correlated noise. This parameter is also
following a Gaussian distribution, with a zero-mean and a covariance
matrix $C_{\eta}$ (i.e. line 10 p5). This approach is not common in
modelling GNSS time series. The standard is based on Williams 2003,
where the author established a stochastic model with a covariance
matrix as sum of identity matrix for the white noise and another
matrix which represents the coloured noise. The coloured noise
covariance matrix is defined differently (see the literature – Bock and
Melgar Reviews of Geophysics 2016, He et al., Journal of geodynamics
2017).}

I have changed my description of the data so that it includes a single
term, $\eta$, that encompasses correlated and uncorrelated noise. I
then break $\eta$ down into correlated and uncorrelated components in
Section 3.1 (eq. 25), where I go into detail about the noise model for
the Pacific Northwest.

@@

\textbf{2- First, the author should give a summary of stochastic noise
modelling in GNSS. I also think that a discussion is needed to relate
to previous models. Why using separated stochastic models? Also I
would comment on the possibility that the (low) spatial correlation
between the parameters in your deterministic model (i.e. secular
velocities) could be absorbed in the estimation of the covariance
matrix $C_{\eta}$.}

In Section 3.1, I acknowledge that noise models for GNSS data have
been thoroughly researched and my FOGM noise model is not a new
concept. I fear that a full summary of stochastic noise modelling
would be too much of a distraction to the reader. However, I have
referred the reader to He et al 2017 for such a review TODO.

We agree that the spatial correlation between paramaters in the
deterministic model can be incorporated into our noise model. We
acknowledge this possibility in Section when we say ``the secular
velocities, $b_i$, are spatially correlated and we could invoke a
tectonic model to form a prior on $b_i$. However, in our application
to the Pacific Northwest, we will be using displacement time series
which are long enough to sufficiently constrain $b_i$ for each
station, avoiding the need to incorporate a prior. Likewise, seasonal
deformation is spatially correlated (Dong et al. 2002; Langbein 2008),
and it may be worth exploring and exploiting such a correlation in a
future study."

@@

\textbf{Now, I am confused When looking at p.8 Section 4.1, the
description of the noise models should be given as a subsection of
Section 2 p.4. I would also improve the literature review. Several
other models have also been discussed (i.e. Montillet et al., 2014
uses a fractional Brownian motion model; ARMA, ARFIMA, GGM or Band
pass noise – see Het et al., 2017).}

The paper is laid out so that Section 2 contains a general description
of the method, and Section 3 contains details that are specific to
each application. The choice of noise model is specific to the
application, and so we feel that it should remain in Section 3. To
help clarify any confusion caused by this layout, we say in Section 2,
``The appropriate noise model may vary depending on the application,
and we hold off on specifying the covariance matrix, $C_{\eta_i}$,
until Section 3.1".

We feel that a full review on stochastic noise models would be a
distraction to the reader, but we refer the reader to He et al 2017
for such as review.

@@

\textbf{Outlier detection:}

\textbf{1- P7 : I would give a longer summary in the introduction about all
the efforts in outlier detection in GNSS time series. The first
paragraph p.7 is not enough.}

In Section 2.2, we have expanded our discussion of previous efforts to
detect outliers in GNSS data.

@@

\textbf{2- Also, the paper would be improved if the proposed outlier detection
method is compared with existing ones such as the Hector software
package (Bos et al., 2013). For example, the comparison between this
method and another one could be done when applying the algorithm in p.
13 (last paragraph below equation 18).}

My outlier detection algorithm is quite similar to what is used in Bos
et al 2013 (and many other papers), in that the outliers are
determined based on the residuals for a best fitting model. The only
difference is that I am fitting the data with a stochastic model
rather than a parametric model consisting of a linear trend and
sinusoids. When using a parametric model, I found that deformation
from slow slip events would sometimes be erroneously identified as an
outlier. The stochastic model used in my outlier detection method is
more flexible and able to describe this transient deformation, and so
only high frequency anomalous observations get identified as outliers.
I have included Figure A1 to demonstrate this.

Of course, a parametric model that includes, for example, B-splines
would be just as effective at detecting outliers without removing
geophysical signal. So I am not suggesting that my outlier detection
method is a significant advancement or even a novel method. Instead, I
am only describing my outlier detection method for the sake of
completeness. The previous version of my paper seems to have put too
much emphasis on the outlier detection method, when the main focus
should remain on the method for estimating transient strain rates. To
help keep my paper focused, I placed the description of the outlier
detection method in the appendix.

@@

\textbf{Results:}

\textbf{The results are generally well explained. However, the use of
different priors and the comparison in the text is sometimes
confusing. Perhaps, a table could summary the different priors and the
main results. It would ease the reading of Section 4.2, 4.3 and then
the discussion in Section 5.}

In the previous draft, I was discussing two different priors: a prior
for the outlier detection algorithm, and a prior used in computing
transient strain rates. I acknowledge that this was convoluted. In the
revised draft, all the details of the outlier detection algorithm are
confined to the appendix. I hope this clarifies any confusion.

@@

\textbf{p.2 line 43 :”too large of an area” ... perhaps “ a very large area”}

Done

\textbf{P5 line 32 : “formal data uncertainty”: can you define it?}

By "formal data uncertainties" I meant "the derived uncertainties that
accompany the GNSS displacement solutions". The text has been
clarified.

### 2017-11-09 RevisedSubmission/hines_hetland_2017_gji.tex
did more

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Consequently, secular crustal
strain rates estimated from GNSS data have been used to constrain
seismic hazard models such as UCERF3 \citep{Field2014}. Transient
crustal strain, which is caused by geophysical phenomena such as slow
slip events (SSEs) or postseismic deformation, is also relevant for
assessing seismic hazard. While transient strain itself is not
damaging, there is a risk that it can trigger major earthquakes
\citep{Roeloffs2006,Freed2001}. Dense networks of continuous GNSS
stations, such as the Plate Boundary Observatory (PBO), make it
possible to identify transient strain with high fidelity. Developing
and improving upon methods for deriving secular and transient strain
from GNSS data is an active area of research.

@@

Most methods for estimating strain rates from GNSS data assume some
parametric form for the deformation signal. The simplest method for
estimating secular strain rates assumes that GNSS derived velocities
can be described with a first-order polynomial (i.e, the deformation
gradients are constant) over some subnetwork of the GNSS stations
\citep[e.g.,][]{Feigl1990,Murray2000}. The components of the strain
rate tensor for each subnetwork are then determined from the least
squares fit to the observations. The assumption that deformation
gradients are spatially uniform is not appropriate when subnetworks
span too large of an area. To help overcome this deficiency,
\citet{Shen1996,Shen2015} used an inverse distance weighting scheme,
in which the estimated strain rate at a point is primarily controlled
by observations at nearby stations. The method of
\citet{Shen1996,Shen2015} can be viewed as a form of local least
squares regression with a first-order polynomial \citep[e.g.,][sec.
6]{Hastie2009}. Other methods for estimating secular strain rates have
parameterized GNSS derived velocities with bi-cubic splines
\citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and
elastostatic Green's functions \citep{Sandwell2016}. The type of basis
functions and the number of degrees of freedom for a parameterization,
which are often chosen subjectively, have a strong influence on the
strain solution. If there are too few degrees of freedom in the
parameterization, then estimated strain rates will be biased and the
uncertainties will be underestimated. On the other hand, if there are
too many degrees of freedom, then there will not be any coherent
features in the estimated strain rates. The methods described by
\citet{Beavan2001} and \citet{Tape2009} also require the user to
specify penalty parameters that control a similar trade-off between
bias and variance in the solution. One could parameterize deformation
with a physically motivated model of interseismic deformation
\citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the
lithospheric rheology and fault geometries are assumed to be known.
Any errors in the assumed physical model could result in biased strain
estimates and underestimated formal uncertainties.

>>>>

Crustal strain rates are fundamentally important quantities for
assessing seismic hazard. Knowing where and how quickly strain is
accumulating gives insight into where we can expect stored elastic
energy to be released seismically. Consequently, secular crustal
strain rates estimated from GNSS data have been used to constrain
seismic hazard models such as UCERF3 \citep{Field2014}. Transient
crustal strain, which is caused by geophysical phenomena such as slow
slip events (SSEs) or postseismic deformation, is also relevant for
assessing seismic hazard. While transient strain itself is not
damaging, there is a risk that it can trigger major earthquakes
\citep{Roeloffs2006,Freed2001}. Dense networks of continuous GNSS
stations, such as the Plate Boundary Observatory (PBO), make it
possible to identify transient strain with high reliability. Developing
and improving upon methods for deriving secular and transient strain
from GNSS data is an active area of research.

@@

Most methods for estimating strain rates from GNSS data assume some
parametric form for the deformation signal. The simplest method for
estimating secular strain rates assumes that GNSS derived velocities
can be described with a first-order polynomial (i.e, the deformation
gradients are constant) over some subnetwork of the GNSS stations
\citep[e.g.,][]{Feigl1990,Murray2000}. The components of the strain
rate tensor for each subnetwork are then determined from the least
squares fit to the observations. The assumption that deformation
gradients are spatially uniform is not appropriate when subnetworks
span a very large area. To help overcome this deficiency,
\citet{Shen1996,Shen2015} used an inverse distance weighting scheme,
in which the estimated strain rate at a point is primarily controlled
by observations at nearby stations. The method of
\citet{Shen1996,Shen2015} can be viewed as a form of local least
squares regression with a first-order polynomial \citep[e.g.,][sec.
6]{Hastie2009}. Other methods for estimating secular strain rates have
parameterized GNSS derived velocities with bi-cubic splines
\citep{Beavan2001}, spherical wavelets \citep{Tape2009}, and
elastostatic Green's functions \citep{Sandwell2016}. The type of basis
functions and the number of degrees of freedom for a parameterization,
which are often chosen subjectively, have a strong influence on the
strain solution. If there are too few degrees of freedom in the
parameterization, then estimated strain rates will be biased and the
uncertainties will be underestimated. On the other hand, if there are
too many degrees of freedom, then there will not be any coherent
features in the estimated strain rates. The methods described by
\citet{Beavan2001} and \citet{Tape2009} also require the user to
specify penalty parameters that control a similar trade-off between
bias and variance in the solution. One could parameterize deformation
with a physically motivated model of interseismic deformation
\citep[e.g.,][]{Meade2005,McCaffrey2007}. In such models the
lithospheric rheology and fault geometries are assumed to be known.
Any errors in the assumed physical model could result in biased strain
estimates and underestimated uncertainties.
